Psychological Needs During Intergroup Contact: Three Experience Sampling Studies

### 2022-11-07 manuscript.tex
Updates from Overleaf
present research str

\section{The Present Research}
The aim of this paper is essentially threefold. First, our study is among the first to test the fundamental tenets of intergroup contact and Allport's conditions in real-life extensive longitudinal data. Second, we test whether the fulfillment of situational psychological needs is meaningfully related to more positive outgroup attitudes following intergroup interactions. And third, we test the role of interaction quality in  the relationship of psychological need fulfillment and outgroup attitudes. Based on this we formulated three main hypotheses:
\begin{enumerate}[label=H\arabic*:]
    \item Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for changes in more favorable outgroup attitudes across extensive longitudinal data.
    \item Based on Allport’s optimal contact conditions, intergroup interactions with equal status, common goals, collaboration, and structural support should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data.
    \item Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data. We also expect situational need fulfillments to work at least as well as Allport's optimal contact conditions in predicting outgroup attitudes.
\end{enumerate}
Beyond these main hypotheses, we have formulated several sub-hypotheses that form the basis for our analysis plan (these sub-hypotheses also include several robustness checks and are available in \appref{app:AppendixHypotheses}).

To test our main hypotheses, we collected three independent sets of extensive longitudinal data (Studies 1–3). The full surveys are available in Supplementary Materials A as well as in our open science repository \citep[including a complete codebook, see][]{KreienkampMasked2022a}. Our most comprehensive study (Study 3) was fully preregistered  \citep[available at][]{KreienkampMasked2021f}. All studies received ethical approval from the University Masked for Peer Review. The fully annotated analyses are available in Supplementary Material B.

\section{Analysis Plan}
\begin{center}
    \noindent\rule{8cm}{0.4pt}\\
    \hl{\textit{New Section:}}\\
    \vspace{-0.25cm}
    \noindent\rule{8cm}{0.4pt}
\end{center}

Given the more complex data structure of the extensive longitudinal data, we opted to test most of our hypotheses using multilevel regression models, where the measurement occasions (level 1) were nested within the participants (level 2). Given that more complex dynamic modeling procedures are not necessary to test these more basic hypotheses, we will focus on the contemporaneous effect of the data (e.g., need fulfillment during the intergroup interaction predicts outgroup attitudes following the interaction). Such an approach is also tolerant to missing data and uneven case numbers within participants. As is common with multilevel analyses, we use a hierarchical modeling approach and always report the final model in-text \citep[][; for the full modeling process see Supplementary Material B]{Bliese2013}.

@@

For our main theory tests we sequentially focus on five sets of models to test and validate our hypotheses:

\begin{enumerate}
    \item \textbf{Contact hypothesis in extensive longitudinal data.} Within the individual studies, we begin by testing the most basic assumption of the intergroup contact hypothesis that outgroup attitudes should be more positive after outgroup interactions but not after non-outgroup interactions. For this we use multi-level regression analyses, predicting outgroup attitudes from outgroup interaction and non-outgroup interaction dummy variables. We also reintroduce the participant means as level two predictors to fully disentangle measurement- (level 1) and participant (level 2) variances.

    Importantly, this initial test of the contact hypothesis in intensive longitudinal data does not consider the influence of interaction quality. This is due to the structural missingness of interaction quality --- where people are only able to report their interaction quality perceptions if they had an interaction. Such structural missingness should not be imputed, which stands in conflict with (multi-level) linear regression analyses that can only be estimated where all variables have valid data (which would only be measurements following interactions). The best way to assess the joint effects of interaction numbers and interaction quality is thus a test on an aggregate level. However, as such an aggregation reduces statistical power to the participant level, we assess the full contact hypothesis in the `embeddedness and stability across studies' section, where we combine the data of the three studies (see below).
    \item \textbf{Core need fulfillment during intergroup contact.} The main proposal of this manuscript has been the assertion that the fulfillment of situation core needs during an interaction will be associated with more positive interaction quality perceptions and, ultimately, more positive outgroup attitudes. Thus, for the main set of analyses, we focus on the reported outgroup interactions only. For each study, we will use multi-level regressions to test the main three assertions of our proposal (mirroring the basic steps of a traditional mediation analysis).
    \begin{enumerate}
        \item Core need fulfillment during outgroup interactions should predict more positive outgroup attitudes. To test this sub-hypothesis, we use a two-level regression analysis of outgroup attitudes, with core need fulfillment as the main predictor variable.
        \item Core need fulfillment during outgroup interactions should also predict higher perceived interaction quality. This sub-hypothesis mirrors the one above so that perceived interaction quality is predicted by core need fulfillment in a multi-level regression analysis.
        \item The effect of core need fulfillment on outgroup attitudes should be reduced when considered together with perceived interaction quality. To test whether perceived interaction quality is indeed the psychological mechanism that explains the variance in outgroup attitudes, we use a multi-level regression analysis of outgroup attitudes with core need fulfillment and perceived interaction quality as the main predictor variables.
    \end{enumerate}
    \item \textbf{Robustness within studies.} To build further confidence in the effect of core need fulfillment during outgroup interactions, we conduct two additional robustness analyses for each study. 
    \begin{enumerate}
        \item To check for the role of alternate psychological needs, we add the fulfillment of self-determination theory needs (i.e., competence, autonomy, and relatedness) to the multi-level regression. We then also compare the model with models that predicts outgroup attitudes from self-determination theory need fulfillments or core need fulfillments only. 
        \item To ensure that the core need fulfillment is outgroup contact specific, we return to the full sample of intensive longitudinal measurements within each study and test whether there is an interaction effect of outgroup contact (vs. no outgroup contact) and core need fulfillment. We expect that the effect of core need fulfillment is specific to outgroup interactions and not merely due to a more need fulfilled life in general.
    \end{enumerate} 
    \item \textbf{Allport's conditions in extensive longitudinal data.} Within the third study we formally measure all of Allport's optimal contact conditions. 
    \begin{enumerate}
        \item We use multi-level regression models to test whether the fulfillment of Allport's conditions in real-life data predicts more positive outgroup attitudes and higher perceived interaction quality, using the same approach as for the core need fulfillment above.
        \item We then compare the effect of Allport's contact conditions with our core need fulfillment by comparing the model fit statistics of the two individual models and by adding both concepts to a joint multi-level regression model, to see whether the two approaches explain the same variance in outgroup attitudes.
    \end{enumerate}
    \item \textbf{Embeddedness and stability across studies.} Within the final set of analyses, we look at the broader picture of our results and leverage the data from all participants to contextualize our results.  
    \begin{enumerate}
        \item We first assess the stability of our main analyses. We use forest plots (including meta analytical estimates) to visualize the direction and effect sizes of our three studies. 
        \item We then combine the data from the three studies to test the basic tenets of intergroup contact hypothesis more broadly within the intensive longitudinal real-life data. As we have mentioned above, assessing the effects of intergroup contacts and perceived interaction quality jointly is difficult within a multi-level regression because quality perceptions are only available when participants had an interaction. To circumvent this issue, we use the data from all participants and aggregate the number of interactions participants had, their average interaction quality perceptions, as well as their average outgroup attitudes. We can then use a linear regression analysis to assess the joint effects of intergroup contacts and interaction quality to predict outgroup attitudes (while controlling for the participants' study). This not only offers a more general test of the contact hypothesis with real-life data, but the aggregated data also mirrors the most common methodology of assessing intergroup contact effects within the cross-sectional literature --- where participants are asked to recall how many interactions they had over a one month period, how positive these interactions were, and what their general attitudes towards the outgroup are. This analysis will thus be able to explore potential discrepancies between our main multi-level analyses, the aggregation of extensive longitudinal data, and traditional participant recall studies.
        \item We, finally, use the qualitative data from the participants' self-identified core needs to contextualize our results. We leverage machine learning to extract a topic model of the free-text entries across the three studies. We describe the extracted topics and themes, and compare them to the need contents usually found and measured within the psychological literature. Full methodological details and visualizations are available in Supplemental Material C.
    \end{enumerate}
\end{enumerate}

@@

>>>>

\section{The Present Research}
The aim of this paper is essentially threefold. First, our study is among the first to test the fundamental tenets of intergroup contact and Allport's conditions in real-life extensive longitudinal data. Second, we test whether the fulfillment of situational psychological needs is meaningfully related to more positive outgroup attitudes following intergroup interactions. And third, we test the role of interaction quality in  the relationship of psychological need fulfillment and outgroup attitudes. Based on our three main aims we developed a set of four main hypotheses, which are formalized into individual sets of models in our analysis plan.

\begin{enumerate}[label=H\arabic*:]
    \item Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for changes in more favorable outgroup attitudes across extensive longitudinal data.
    \begin{enumerate}[label=H1\alph*:, leftmargin=\parindent]
        \item Outgroup attitudes should be more positive after an intergroup interaction compared to a non-outgroup interaction.
        \item Participants with more intergroup interactions should have more favorable outgroup attitudes
        \item Participants with more intergroup interactions should have more favorable outgroup attitudes depending on the average interaction quality.
    \end{enumerate}
    \item Based on Allport’s optimal contact conditions, intergroup interactions with equal status, common goals, collaboration, and structural support should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data.
    \begin{enumerate}[label=H2\alph*:, leftmargin=\parindent]
        \item Higher fulfillment of Allport’s conditions during out-group interactions should predict more positive outgroup attitudes.
        \item Higher fulfillment of Allport's conditions during outgroup interactions should also predict higher perceived interaction quality.
        \item The effect of higher fulfillment of Allport's conditions on outgroup attitudes should be reduced when considered together with perceived interaction quality.
    \end{enumerate}
    \item Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data. We also expect situational need fulfillments to work at least as well as Allport's optimal contact conditions in predicting outgroup attitudes.
    \begin{enumerate}[label=H3\alph*:, leftmargin=\parindent]
        \item Core need fulfillment during outgroup interactions should predict more positive outgroup attitudes.
        \item Core need fulfillment during outgroup interactions should also predict higher perceived interaction quality.
        \item The effect of core need fulfillment on outgroup attitudes should be reduced when considered together with perceived interaction quality.
    \end{enumerate}
    \item stable and robust (To build further confidence in our results, we assessed two additional models that might offer alternative explanations).
    \begin{enumerate}[label=H4\alph*:, leftmargin=\parindent]
        \item The effect of key need fulfillment on outgroup attitudes should persist even when taking other fundamental psychological needs into account. Thus, the effect of key need fulfillment on outgroup attitudes should remain strong even after controlling for autonomy, competence, and relatedness fulfillment during the interaction (cf., self-determination theory).
        \item The effect of key need fulfillment on outgroup attitudes should be specific to intergroup interactions and not be due to need fulfillment in general. Thus, the effect of key need fulfillment on outgroup attitudes should be stronger for intergroup interactions than for ingroup interactions.
        \item The effects of our main hypotheses and robustness analyses should be consistent across studies.
    \end{enumerate}
\end{enumerate}
Beyond these main hypotheses, we have formulated several sub-hypotheses that form the basis for our analysis plan (for the full sub-hypotheses and analysis plan see \appref{app:AppendixHypotheses}).

To test our main hypotheses, we collected three independent sets of extensive longitudinal data (Studies 1–3). The full surveys are available in Supplementary Materials A as well as in our open science repository \citep[including a complete codebook, see][]{KreienkampMasked2022a}. Our most comprehensive study (Study 3) was fully preregistered  \citep[available at][]{KreienkampMasked2021f}. All studies received ethical approval from University Masked for Peer Review. The fully annotated analyses are available in Supplementary Material A.

Given the more complex data structure of the extensive longitudinal data, we opted to test most of our hypotheses using multilevel regression models, where the measurement occasions (level 1) were nested within the participants (level 2). Such an approach is also tolerant to missing data and uneven case numbers within participants. As is common with multilevel analyses, we use a hierarchical modeling approach and always report the final model in-text \citep[][; for the full modeling process see Supplementary Material B]{Bliese2013}.

@@

@@

\subsection{Stability and Conclusion}
As with the main analyses, we use a forest plot and the associated meta-analytic estimates to assess the stability of the results across the three studies. The forest plot shows that the core need fulfillment is also consistently outgroup contact specific (\fgrref[-A]{fig:Robustness}) and remains a meaningful predictor of outgroup attitudes even when controlling for traditional psychological need fulfillments (Figure \fgrref[-B]{fig:Robustness}). In sum, we thus find that across all three studies, psychological need fulfillment remained a robust and flexible predictor of positive outgroup attitudes.
